---
title: " "
author: "January Weiner"
date: "`r Sys.Date()`"
output:
  ioslides_presentation:
    widescreen: true
    smaller: true
    css: "../files/style.css"
---

```{r,echo=FALSE}
## Set default options for the knitr RMD processing
knitr::opts_chunk$set(echo=TRUE,eval=FALSE,warning=FALSE,message=FALSE,fig.width=5,fig.height=5,cache=FALSE,autodep=TRUE, results="hide")
library(tidyverse)
library(readxl)
library(beeswarm)
```



# Mapping

## How to understand mapping

In R, functions can be passed as arguments.

```r
my_lapply <- function(x, func) {
  n <- length(x)
  ret <- list()
  for(i in 1:n) {
    ret[[i]] <- func(x[[i]])
  }
  names(ret) <- names(x)
  return(ret)
}
```

## Mapping in R / in R + tidyverse

```{r}
tmp <- list(a=1, b=2, c=1:10)

lapply(tmp, length)
lapply(tmp, function(x) paste(x, collapse=", "))
```


## Mapping in R / in R + tidyverse

```{r}
map(tmp, ~ paste(., collapse=", "))

## type conversion is always specified
map_chr(tmp, ~ paste(., collapse=", "))
```

## Why both worlds?

 * The \*pply functions are still widely used, so you will see them in code
 * sometimes using tidyverse does not really give you a bonus (90% of the
   cases a regular lapply will do)

## Why not use `group_by` and `summarise`?

Thes functions work on data frames – lists of vectors of *equal* length.
They don't work on lists of vectors which have different lengths!

On the other hand, \*pply family and the purrr functions work also on data
frames (because data frames are lists!).

## lapply & sapply

`lapply` takes a list and returns a list, always.

`sapply` takes a list and tries to simplify it to a relevant object (data
frame, vector etc.). Which sometimes fails.

Compare:

```{r,eval=FALSE}
lapply(starwars$name, nchar)
sapply(starwars$name, nchar)
```

## `replicate`

This is a special form of `sapply` which simply calls a function `n` times
and returns the results in a list. *Note that it uses an expression and not
a function as a second argument.*

```r
## create 10 vectors with random numbers
replicate(10, rnorm(10), simplify=FALSE)
```

## General form

```r
xxxapply(data, ..., function)
```

Where `function` can be a closure (anonymous function), defined in place:

```r
## replace spaces by underscores
sapply(starwars$name, 
  function(x) 
      gsub(" ", "_", starwars$name))
```


## apply

`apply` is for data frames and matrices (two-dimensional objects).
The second argument determines whether it works row-wise (1) or column-wise
(2):


```r
## matrix of random numbers, 3 x 20
mtx <- matrix(rnorm(60), ncol=3)

apply(mtx, 1, sum)
apply(mtx, 2, sum)
```

## There is a simpler way

Don't forget following convenience functions:

 * `colSums`, `colMeans`
 * `rowSums`, `rowMeans`

(However, there is no colSD or rowSD)

## tapply

`tapply`  applies a function to vector split by levels of a factor. For
example, we might want to know what is the mean mpg for cars with different
number of cylinders:

```{r}
tapply(mtcars$mpg, mtcars$cyl, mean)
```

This is a really important function.

## Exercise 3/1

# `purrr`

## `map` and co.

`map` works very much like `sapply(..., simplify=F)`, and its variants
explicitely try to cast the result in an appriopriate format.

 * `map_lgl` -- logical vector
 * `map_chr` -- character vector
 * `map_int` -- integer vector
 * `map_dbl` -- double (floating point) vector

## The weird syntax of `map`

To make the map functions more terse, map family adopted a particular way
of formulating custom functions. `.` (dot) is automatically used as
argument, and instead of writing `function(x) ...` you write `~ ...`:

 * sapply:

        ## return 1, 2, ... 10 random numbers
        sapply(1:10, function(x) rnorm(x))

 * map:

        map(1:10, ~ rnorm(.))

## Formula notation in dplyr

The same notation can be used also in dplyr with the `mutate_all` and
`rename_all` functions:

```r
## Replace "none" by NA 
starwars %>% mutate_all(~ gsub("none", NA, .))

## Modify all variable names
starwars %>% rename_all(~ paste0("Var_", .))
```

## `map_dfr` and `map_dfc`

These two create a data frame. If your function returns each time a data
frame, map will join these – either by rows (`map_dfr`) or by
columns (`map_dfc`).

```r
## create a list with 10 random vectors
rv <- map(1:10, ~ rnorm(10))
names(rv) <- LETTERS[1:10]

## calculate the mean and sd
map_dfr(rv, ~ data.frame(mean=mean(.), sd=sd(.)))
map_dfc(rv, ~ c(mean=mean(.), sd=sd(.)))
```

## `map2` et co.

Sometimes you want to go over two or more vectors at the same time. 
Instead of `.` you have two automatic variable names, `.x` and `.y`.

```r
## add number of vehicles to each name
sw_names <- starwars$name
sw_vehcls <- starwars$vehicles
map2_chr(sw_names, sw_vehcls, 
  ~ paste0(.x, "_", length(.y)))
```

## `pmap` and co.

`pmap` uses a list of vectors of identical length and passes all its
columns as `...` to a function.

```r
data <- list(a=1:3, b=rnorm(3), c=3:1)
pmap(data, ~ sum(...))
```


## `pmap` and co.

Remember: data.frame is also a list! So you
can apply pmap to a data frame and process each row.

```r
starwars %>% 
  select(name, hair_color, eye_color) %>% 
  mutate_all(~ gsub("none", NA, .)) %>%
  drop_na() %>%
  pmap_chr(~ sprintf("%s has %s hair and %s eyes", ...))
```

## Exercise 3/2

# Regular expressions

## regexp: regular expressions

regexp is for matching patterns to strings. A basic unit of regexp is an
atom, which can be followed by a modifier. Examples:

 * `a` – (atom) match the literal "a" (and not A!)
 * `a*` – (atom + modifier) match "a" any number of times (including zero!)
 * `a+` – (atom + modifier) match "a" at least once
 * `a?` – (atom + modifier) match "a" zero or once
 * `.` – (atom) match anything
 * `.*` – (atom + modifier) match anything any number of times
 * `[...]` – (atom) match any letter between the brackets
 * `[^...]` – (atom) match anything *but* a letter between the brackets
 

## Other important modifiers

  * `^`: if at the beginning of a regexp, then it matches only the
    beginning of the string, so `^oo` matches "oo" but not "Atchoo".
  * `$`: same, but at the end of a regexp: `oo$` matches "poo" but not
    "Winnie the Pooh"
  * `^oo$` matches "oo" and nothing else

## how to match a modifier?

How do you match a star, or a question mark?

Prepend it with *two* backslashes:

    x <- c("abcde", "abce","abcd?e")

    # matches "abcde" and "abce":
    grep("abcd?e", x)

    # matches "abcd?e":
    grep("abcd\\?e")

## Why two backslashes

Backslash in R is *always* for creating "special" characters. For example,
`\n` means a new line. So to actually get a backslash, you need to precede
it with a backslash: `\\`. In other languages it may be different or depend
on the context.

## What can we use it for

 * Searching through a vector: with `grep()`  or `grepl()`
 * Substituting in a vector: with `sub()`  or `gsub()`
 * Searching within a string: with `regexpr()` or `gregexpr()`

## The coolest part: groups

You can easily extract matched fragments from strings:

```{r}
input <- "name:January last_name:Weiner"

gsub("name:(.*) last_name:(.*)", "\\2 \\1", input)
```

## Exercise 3/3

